Progress Report

### 2023-10-02 Docs/ProgressReport.Rmd
fit debias model in the fast way

df <- read_rds(here("Data/nhanes_bi.rds"))
# load(here("Data/ApplOutput_fGFPCA.RData"))

@@

@@

```{r}
# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))
```

### Bin Data

@@

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# head(df)

df %>% 
  filter(SEQN %in% rand_id) %>%
  group_by(SEQN, bin) %>%
  summarise(num = sum(Z)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")

df <- df %>% rename(id = SEQN, Y=Z)
```
### Data split

```{r}
id_vec <- unique(df$id)
train_id <- sample(id_vec, size = 0.8*N, replace = FALSE) # 7010 subjects for training
test_id <- setdiff(id_vec, train_id) # 1753 subjects for testing
```

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction

@@

```{r, cache=TRUE}
# fit model on the training set
df_train <- df %>% filter(id %in% train_id)
train_bin_lst <- split(df_train, f = df_train$bin)

@@

```{r}
df_est_latent <- bind_rows(df_est_latent) %>% 
  select(-sind, -Y) %>% distinct(.) 
# example estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r, message=FALSE}

df %>% 
  filter(id %in% sample(train_id, size = 4)) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  facet_wrap(~id)+
  labs(x="Time", y = "Estimated latent function (original scale")
```

@@

```{r}
mat_est_unique <- matrix(df_est_latent$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) # row index subject, column binned time
#dim(mat_est_unique)

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model

# dim(fpca_mod$efunctions) # 27 eigenfunctions total
# fpca_mod$evalues
K <- 4
```

@@

1. Interpolate the eigenfunctions with B-spline basis back to the original grid
2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction
- How are the eigenvalues biased? 
- It looks like the FPCA step now estimates mean and eigenfunctions, while each eigenfunction serves as a random slope. Can we fit such model with an offset of mean/intercept function?

First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. The model failed because of vector memory exhaustion. Shall I: 
- Decrease the density of grid (increase bin width)?
- Decrease sample size? 
- Increase memory (re-set to 100GB). But still took a lot of time (more than one hour). The model actually did not finish fitting.

```{r, eval=FALSE, class.source = "fold-show"}
df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
colnames(df_phi) <- c("bin", paste0("phi", 1:4))

df_train <- df_train %>% left_join(df_phi, by = "bin")
df_train$id <- as.factor(df_train$id)

# usethis::edit_r_environ()
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(bin, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=df_train, method = "fREML")
t2 <- Sys.time()
t_debias <- t2=t1

Then I would like to interpolate eigenfunctions using B-spline basis functions:

>>>>

# df <- read_rds(here("Data/nhanes_bi.rds"))
# load(here("Data/ApplOutput_fGFPCA.RData"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))

@@

# Simulation

I am now on the way to add the debias step to the whole algorithm. Just so that I have a set of "true parameters" to comparision, I am gonna use the simulation dataset for this part.

## Generation mechanism

- I am gonna generate only one dataset as an example
- 500 individuals each with 1000 measures

\[\begin{aligned}
Y_i(t) & \sim Bernoulli(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}\]

where:

- $t$ is equal-spaced on $[0, 1]$
- $f_0(t)=0$
- $\xi_k \sim N(0, \lambda_k)$, and $\lambda_k = 1, 0.5, 0.25, 0.125$ for k = 1, 2, 3, 4 respectively.

```{r gen_data, class.source='show'}
N <- 500 # sample size
J <- 1000 # number of observation points

t = seq(0,1,len=J) # observations points

# mean function
f_0 <- function(s) 0

#eigenfunctions 
K <- 4 # number of eigenfunctions
phi <- sqrt(2)*cbind(sin(2*pi*t),cos(2*pi*t),
                     sin(4*pi*t),cos(4*pi*t))

# eigenvalues
lambda = 0.5^(0:(K-1))

# generated data 
## score
xi <- matrix(rnorm(N*K),N,K)
xi <- xi %*% diag(sqrt(lambda))

# subject-specific random effect
b_i <- xi %*% t(phi); # of size N by J

# latent gaussian function
eta_i <- t(vapply(1:N, function(x){
    f_0(t) + b_i[x,]
  }, numeric(J)))

# outcome binary function
Y_i <- matrix(rbinom(N*J, size=1, prob=plogis(eta_i)), 
                N, J, byrow=FALSE)

# format into dataframe
# id = subject identifier (factor variable)
# sind = numeric value corresponding to the observed functional domain
# Y = functional response (binary)
# sind_inx = numeric value associated with order of "sind"
#            this is not necessary, but may be of convenience when you implement the method
df <- data.frame(id = factor(rep(1:N, each=J)),
                 t = rep(t, N), 
                 Y = as.vector(t(Y_i)),
                 eta_i = as.vector(t(eta_i)),
                 sind = rep(1:J, N))
# visualization
rand_id <- sample(N, size = 4)
df %>% filter(id %in% rand_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y))+
  geom_line(aes(x=sind, y=plogis(eta_i)), col = "red")+
  #geom_line(aes(x=sind_inx, y=eta_i), col = "blue")+
  facet_wrap(~id)
```

@@

### Bin ata

@@

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# unique(df$bin)

df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~id)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```

## Data split

- 400 subjects for training, 100 for out-of-sample prediction

```{r}
train_id <- sample(unique(df$id), size = N*0.8)
test_id <- setdiff(unique(df$id), train_id)

train_df <- df %>% filter(id %in% train_id)
test_df <- df %>% filter(id %in% test_id)
```

@@

```{r, cache=TRUE}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

@@

```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  mutate(eta_i = exp(eta_i)/(1+exp(eta_i))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_line(aes(x=sind, y=eta_i, group = id, col = "true"))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

@@

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)

mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=length(train_id), 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time
# dim(mat_est_unique)

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T)
t2 <- Sys.time()
t_fpca <- t2-t1 # 3.21 seconds to fit fPCA model
```

@@

1. Project the eigenfunctions with B-spline basis back to the original grid
2. Debias the eigenvalues with GLMM (mgcv::bam, mgcm::gamm or gamm4:gamm) to be used for Laplace approximation

- The bias was caused by the misspecification of latent process (assume constant effect over smooth function). 
- What needs re-evaluation are eigenvalues (variance estimates). They will be used in Laplace Approximation for out-of-sample prediction. 
- We do not need to debias individual scores because the scores for training sample will not be used for out-of-sample prediction

#### Debias with GLMM model

First I'd like to try to fit the debias GLMM model on the full training set with un-interpolated eigenfunctions. I am using mgcv::bam and setting method = "fREML" and discrete = TRUE to speed up computation

```{r}
df_phi <- data.frame(bin=mid, fpca_mod$efunctions[, 1:4])
colnames(df_phi) <- c("bin", paste0("phi", 1:4))

# train_df$bin <- as.numeric(as.character(train_df$bin))
train_df <- train_df %>% left_join(df_phi, by = "bin")
train_df$id <- as.factor(train_df$id)
```

- Here, what if I replace "bin" with "t"?

```{r, class.source = "fold-show"}
# usethis::edit_r_environ()
t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(t, bs="cr", k=10)+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, data=train_df %>% filter(bin==sind), 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
t_debias <- t2-t1
```

- And now I try to extract the variance of scores (debiased eigenvalues). 
- Use the conclusion that connects smoothing parameters to variance

$$\hat{p}_k=1/\hat{\lambda}_k^2$$

```{r}
fpca_mod$evalues
new_lambda <- 1/debias_glmm$sp[2:5]
new_lambda/lambda
fpca_mod$evalues[1:4]/new_lambda
```

- Looks like the re-evaluated variance is still inflated from the true eigenvalues, but indeed doubled the result of FPCA.

#### Projection

Then I would like to project eigenfunctions using B-spline basis functions:
